Make interactive loan word maps

### 2025-06-04 analysis/paper/loan word maps.qmd
static maps

## Retrieve and clean data
In order to have location data, we read in the African languages, their glottoglog code, and their location from the glottolog database (@hammarstromGlottolog2024).

@@

### Add further layers to the map
Get the archaeological data layers
```{r}
# Create a vector of file paths
points <- here::here("analysis", "data", "raw_data", c("gold_points.csv", 
                                                     "fonio_points.csv",
                                                     "millet_points.csv",
                                                     "salt_points.csv",
                                                     "iron_points.csv"))

@@

function to add further layers
```{r}
add_sf_to_leaflet_map <- function(leaflet_map, new_data_sf, label_column) {

@@

## Networks
Next, we will create networks from this data, in order to be able to analyse and visualise connections.The dataset of languages is the nodelist, the edges are made on the basis of shared forms. First, the edge list is made. Fir this, we need more detailed classifications than for the maps above, so we need to read the data in again without simplifying the classifications.

@@

now, we read this into tidygraph network objects. First we take the language list and convert it into a usable edge list, and then make both into a network.
```{r}
# Convert langloc into a tibble with latitude and longitude fields
node_list <- langloc |>
  dplyr::mutate(
    latitude = sf::st_coordinates(geometry)[, 2],
    longitude = sf::st_coordinates(geometry)[, 1]
  ) |>
  tibble::as_tibble() |> 
  dplyr::select(glott_code, LanName, latitude, longitude) |>

@@

### Fonio
```{r }
plot_fr_graph(graphs$fonio, title = "fonio")
```

@@

### Create a cost variable
This combines weight and geographic distance.
```{r}
graph <- graphs$gold

@@

### Via edge betweenness
Goal: Find edges that are most frequently used in shortest paths — "key trade routes"
```{r}
graph <- graph |> 
  tidygraph::activate(edges) |> 
  tidygraph::mutate(inv_cost = ifelse(cost == 0, Inf, 1 / cost)) |> # cost variable needs to be inverted
  tidygraph::mutate(edge_betweenness = centrality_edge_betweenness(weights = inv_cost))

@@

    Goal: Find the most efficient trade routes considering both weight (stronger trade) and distance (shorter path).

@@

    Geographic distance can be calculated using the Haversine formula or geosphere::distHaversine().

@@

    Goal: Find edges that are most frequently used in shortest paths — "key trade routes".

@@

    You’ll need a cost column (e.g., inverse of weight or distance/weight) as edge weights for the centrality calculation.

@@

    Goal: Assess trade capacity or bottlenecks.

@@

        This identifies routes with the greatest potential trade capacity.

@@

    Goal: Extract a simplified network showing the most critical, efficient paths covering the entire network.

@@

    This highlights low-cost, high-value trade connections.

@@

    Goal: Find routes that link otherwise separated regions.

@@

        These are often interregional trade routes.

@@

    Goal: Visually identify trade corridors.

@@

        Highlight high-weight, long-distance edges or edges with high betweenness.

>>>>

## Retrieve and clean data

In order to have location data, we read in the African languages, their glottoglog code, and their location from the glottolog database (@hammarstromGlottolog2024).

@@

### Add further layers to the map

Get the archaeological data layers

```{r}
# Create a vector of file paths
points <- here::here("analysis", "data", "raw_data", c("gold_points.csv", 
                                                     "fonio_points.csv",
                                                     "millet_points.csv",
                                                     "salt_points.csv",
                                                     "iron_points.csv"))

@@

function to add further layers

```{r}
add_sf_to_leaflet_map <- function(leaflet_map, new_data_sf, label_column) {

@@

## Make static maps

```{r}
library(rnaturalearth)

# Load coastline and river data
coastline <- ne_coastline(scale = "medium", returnclass = "sf")
rivers <- ne_download(scale = "medium", type = "rivers_lake_centerlines", category = "physical", returnclass = "sf")
```

```{r}
plot_language_map <- function(data, coastline, rivers, bbox_padding = 0.1) {
  # Ensure consistent CRS across all inputs
  common_crs <- sf::st_crs(coastline)  # Use the CRS of the coastline as reference
  data <- sf::st_transform(data, crs = common_crs)
  rivers <- sf::st_transform(rivers, crs = common_crs)

  # Compute the bounding box of the data points
  bbox <- sf::st_bbox(data)
  # Extract numeric values for xmin, xmax, ymin, ymax to ensure clean handling
  bbox_xmin <- as.numeric(bbox["xmin"])
  bbox_xmax <- as.numeric(bbox["xmax"])
  bbox_ymin <- as.numeric(bbox["ymin"])
  bbox_ymax <- as.numeric(bbox["ymax"])

  # Compute expanded bounding box
  expanded_bbox <- sf::st_bbox(
    c(
      xmin = bbox_xmin - bbox_padding * (bbox_xmax - bbox_xmin),
      xmax = bbox_xmax + bbox_padding * (bbox_xmax - bbox_xmin),
      ymin = bbox_ymin - bbox_padding * (bbox_ymax - bbox_ymin),
      ymax = bbox_ymax + bbox_padding * (bbox_ymax - bbox_ymin)
    ),
    crs = sf::st_crs(data)  # Use CRS from input data
  )
  # Convert bbox to polygon as sf object
  expanded_bbox_sf <- sf::st_as_sfc(expanded_bbox)

  # Crop coastline and rivers to match the bounding box
  coastline_cropped <- sf::st_intersection(coastline, expanded_bbox_sf)
  rivers_cropped <- sf::st_intersection(rivers, expanded_bbox_sf)

  # Extract coordinates and drop geometry
  data <- data |>
    dplyr::mutate(
      lon = sf::st_coordinates(geometry)[, 1],
      lat = sf::st_coordinates(geometry)[, 2]
    ) |>
    sf::st_drop_geometry()

  # Count occurrences of each category and order
  category_order <- data |>
    dplyr::count(style, name = "freq") |>
    dplyr::arrange(dplyr::desc(freq)) |>
    dplyr::mutate(color = dplyr::case_when(
      dplyr::row_number() <= 12 ~ rcartocolor::carto_pal(n = 12, name = "Bold")[dplyr::row_number()], 
      TRUE ~ "#3b3b3b"
    ))

  # Merge color mapping back to the data
  data <- dplyr::left_join(data, category_order, by = "style")

  # Plot the map
  ggplot2::ggplot() +
    ggplot2::geom_sf(data = coastline_cropped, color = "black", linewidth = 0.5) +  # Cropped coastline
    ggplot2::geom_sf(data = rivers_cropped, color = "gray", linewidth = 0.3) +      # Cropped rivers
    ggplot2::geom_text(
      data = data,
      ggplot2::aes(x = lon, y = lat, label = form, color = style),
      position = ggplot2::position_jitter(width = 1.5, height = 1.5), 
      size = 3
    ) +  # Labels with jitter
    ggplot2::scale_color_manual(values = setNames(category_order$color, category_order$style)) +  # Custom palette with dark gray fallback
    ggplot2::coord_sf(
      xlim = c(expanded_bbox["xmin"], expanded_bbox["xmax"]),
      ylim = c(expanded_bbox["ymin"], expanded_bbox["ymax"]),
      expand = FALSE
    ) +  # Map extent controlled by `expanded_bbox`
    ggplot2::theme_void() +
    ggplot2::theme(legend.position = "none")
}
```

```{r}
plots <- purrr::map(data_list, ~plot_language_map(.x, coastline, rivers, bbox_padding = 0.15))
```

```{r}
plot_language_map(data_list$gold, coastline, rivers, bbox_padding = 0.05)
```

## Networks

Next, we will create networks from this data, in order to be able to analyse and visualise connections.The dataset of languages is the nodelist, the edges are made on the basis of shared forms. First, the edge list is made. Fir this, we need more detailed classifications than for the maps above, so we need to read the data in again without simplifying the classifications.

@@

now, we read this into tidygraph network objects. First we take the language list and convert it into a usable edge list, and then make both into a network.

```{r}
# Convert langloc into a tibble with latitude and longitude fields
node_list <- langloc |>
  dplyr::mutate(
    latitude = sf::st_coordinates(geometry)[, 2],
    longitude = sf::st_coordinates(geometry)[, 1]
  ) |>
  tibble::as_tibble() |> 
  dplyr::select(glott_code, LanName, latitude, longitude) |>

@@

### Fonio

```{r }
plot_fr_graph(graphs$fonio, title = "fonio")
```

@@

### Create a cost variable

This combines weight and geographic distance.

```{r}
graph <- graphs$gold

@@

### Via edge betweenness

Goal: Find edges that are most frequently used in shortest paths --- "key trade routes"

```{r}
graph <- graph |> 
  tidygraph::activate(edges) |> 
  tidygraph::mutate(inv_cost = ifelse(cost == 0, Inf, 1 / cost)) |> # cost variable needs to be inverted
  tidygraph::mutate(edge_betweenness = centrality_edge_betweenness(weights = inv_cost))

@@

```         
Goal: Find the most efficient trade routes considering both weight (stronger trade) and distance (shorter path).

@@

Geographic distance can be calculated using the Haversine formula or geosphere::distHaversine().
```

@@

```         
Goal: Find edges that are most frequently used in shortest paths — "key trade routes".

@@

You’ll need a cost column (e.g., inverse of weight or distance/weight) as edge weights for the centrality calculation.
```

@@

```         
Goal: Assess trade capacity or bottlenecks.

@@

    This identifies routes with the greatest potential trade capacity.
```

@@

```         
Goal: Extract a simplified network showing the most critical, efficient paths covering the entire network.

@@

This highlights low-cost, high-value trade connections.
```

@@

```         
Goal: Find routes that link otherwise separated regions.

@@

    These are often interregional trade routes.
```

@@

```         
Goal: Visually identify trade corridors.

@@

    Highlight high-weight, long-distance edges or edges with high betweenness.
```
